---
title: Outlier Detection in Water Quality Variables Utilising the Temporal Correlation
subtitle: Importance of having multiple sensors close to each other
author: Puwasala Gamakumara
date: \today
classoption: compress
toc: false
header-includes:
  \usepackage{amsmath, nccmath, graphicx}
  \usepackage{bm}
  \usepackage{mathpazo}
output: 
  binb::monash:
    fig_width: 8
    fig_height: 6
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message=FALSE, 
                      warning=FALSE, 
                      cache=TRUE, 
                      dev.args=list(bg=grey(0.9), 
                                    pointsize=11))
library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
```

## Anomaly detection framework

```{r framework, echo=FALSE, fig.width=6, fig.height=4}

knitr::include_graphics('plots/Framework.pdf')

```

<!-- ## Why multiple sensors? -->

<!-- ### Outlier detection -->
<!-- - Single sensor may not be able to detect certain type of anomalies -->
<!-- - If the sensors are malfunctioning, then the covariates or lagged values of them may be incorrect -->

<!-- ### Importance of replication -->
<!-- - Uncertain measurements or unreliable equipment -->
<!-- - It's important to have replications to measure uncertainty and bias -->

<!-- --- -->

<!-- ### Other places where use multiple sensors  -->

<!-- - NEON uses two sensors (about 200m apart) to measure water quality variables -->

<!-- - They initially design this to calculate stream metabolism -->

<!-- - They show that having multiple sensors is important for anomaly detection -->


<!-- --- -->




<!-- - We model the downstream variable using related upstream variables -->

<!-- - It's important to calculate the lag time between the two sensors -->

<!-- - Using cross-correlation methods to calculate lag time -->

<!-- - We then use the residuals from this model to detect outliers -->

<!-- - Finally we compare that with the residuals from a model that uses only the variables from downstream sensor -->

---


### Data

- Pringle Creek - Texas, USA

- Two sensor locations ($200m$ apart)

- **Variables available** - Turbidity, Conductivity, Level and Temperature

- **Time span** - 01-10-2019 to 31-12-2019

- **Frequency** - 5 minute intervals

---

### Time plots

```{r timeplots, echo=FALSE, fig.width=6, fig.height=4}

knitr::include_graphics('plots/Timeplots.pdf')

```


---

### Turbidity downstream

```{r turbplot, echo=FALSE, fig.width=6, fig.height=4}

knitr::include_graphics('plots/Turbidity_downstream.pdf')

```

---

### Model - 1

\begin{gather}
\scalebox{0.75}{$
   \begin{align*}
    log(turbidity\_downstream_t) \sim  
s_1(log(turbidity\_upstream_{t-d_t}))\\ + s_2(conductance\_upstream_{t-d_t}) + 
s_3(level\_upstream_{t-d_t})\\ + 
s_4(temperature\_upstream_{t-d_t}) + \epsilon_t
   \end{align*}$}
\end{gather}

### Model - 2

\begin{gather}
\scalebox{0.75}{$
   \begin{align*}
    log(turbidity\_downstream_t) \sim  
s_2(conductance\_dowstream_{t-1})\\ + 
s_3(level\_downstream_{t-1}) + 
s_4(temperature\_downstream_{t-1}) \\ + \epsilon_t
   \end{align*}$}
\end{gather}

### Model - 3

\begin{gather}
\scalebox{0.75}{$
   \begin{align*}
    log(turbidity\_downstream_t) \sim 
    s_1(log(turbidity\_upstream_{t-d_t}))\\ + s_2(conductance\_upstream_{t-d_t}) + 
s_3(level\_upstream_{t-d_t})\\ + 
s_4(temperature\_upstream_{t-d_t}) + 
s_5(conductance\_downstream_{t-1})\\ + 
s_6(level\_downstream_{t-1}) + 
s_7(temperature\_downstream_{t-1}) \\ + \epsilon_t
   \end{align*}$}
\end{gather}

---

### Residuals from the three models

```{r vis_resid, echo=FALSE, fig.width=6, fig.height=4}

knitr::include_graphics('plots/plot_residuals.pdf')

```

- We use Peak-Over-Threshold method to estimate the outlier threshold

---

## Results

```{r echo=FALSE}
library(kableExtra)

load(here::here("ARCLP-DES-Catchup","Meeting-3", "plots", "confussion_matrix.rda"))

confussion_matrix %>% 
  kable() %>% 
  kable_styling(font_size = 8) %>% 
  row_spec(0,bold=TRUE)
  
```

```{r conf_matrix, echo=FALSE, fig.width=5, fig.height=3, out.width="80%"}

knitr::include_graphics('plots/outliers_turbidity_model3.pdf')

```

<!-- ## Articles and R packages -->

<!-- ### Articles -->

<!-- - Conditional Normalisation in Time Series Analysis -->
<!-- - Anomaly Detection in River Networks -->

<!-- ### R packages -->

<!-- - **`conduits`** (Conditional UI for Time Series normalisation) - https://github.com/PuwasalaG/conduits -->

<!-- - **`dori`** (Data for Outlier Detection in River networks) - https://github.com/pridiltal/dori -->





